Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2018-04-29 manuscript/manuscript.tex
updated Sec III

The optimal statistic is constructed from the autocovariance and cross-covariance matrices 
${\bf{P}}_a$ and ${\bf{S}}_{ab}$,
\begin{eqnarray}
	{\bf{P}}_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	{\bf{S}}_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,.
\end{eqnarray}
where $\res_a$ is a vector of the residuals of the $a$th pulsar in the PTA. 
%In the frequency domain, these can be written as
%\begin{eqnarray}
%	\left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle_{ij} &=& {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_I(f) \right] {\bf{R}}_I^T \,, \\
%	\left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{ij} &=& \chi_{IJ} {\bf{R}}_I \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_J^T \,,
%\end{eqnarray}
%where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
%two observations, 
%$\mathbf{R}$ is the timing model fitting matrix, 
%$\chi_{IJ}$ is the overlap reduction function (ORF) that describes the spatial correlation between two pulsars' residuals, 
%$P_\mathrm{gw}(f)$ is the GW power spectrum, 
%and $P_I(f)$ is the power spectrum for the $I$th pulsar. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) } \,,
\end{equation}
where $\tilde{{\bf{S}}}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{{\bf{S}}}_{ab} = {\bf{S}}_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
If $\Agw=0$, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{-1/2} \,.
\end{equation}
For a measured value of $\hat{A}^2$, 
the significance of $\hat{A}^2 \neq 0$ is given by the signal-to-noise ratio (SNR) 
\begin{equation}
	\rho = \frac{\hat{A}^2}{\sigma_0} = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2}} \,.
\end{equation}
The expectation value of the SNR, averaged over all realizations of the stochastic background, is
\begin{equation}
	\left\langle \rho \right\rangle = \Agw^2 \left[ \sum_{ab} \Tr\left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2} \,.
\end{equation}

@@

We used these three techniques to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets. 
These consisted of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}.
with the observation times increased so that the longest-observed pulsars 
had TOAs spanning over 11 years. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details).
\begin{table}[tb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs. \sv{finish filling in this table}}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & 11.0 & & -13.93 & 3.56 \\
		J0613$-$0200 & 11.0 &  & -13.14 & 1.22 \\
		J1012+5307 & 11.0 &  & -12.79 & 1.51 \\
		J1024$-$0719 & 6.0 & & $-$ & $-$ \\
		J1455$-$3330 & 11.0 & & $-$ & $-$ \\
		J1600$-$3053 & 8.0 & & $-$ & $-$ \\
		J1614$-$2230 & 7.0 & & $-$ & $-$ \\
		J1640+2224 & 11.0 & & $-$ & $-$ \\
		J1713+0747 & 11.0 & & -14.14 & 1.58 \\
		J1741+1351 & 6.0 & & $-$ & $-$ \\
		J1744$-$1134 & 11.0 & & $-$ & $-$ \\
		B1855+09 & 11.0 & & -13.75 & 3.54 \\
		J1853+1303 & 7.0 & & $-$ & $-$ \\
		J1909$-$3744 & 11.0 &  & -13.84 & 1.74  \\
		J1918$-$0642 & 11.0 & & $-$ & $-$ \\
		J2010$-$1323 & 7.0 & & $-$ & $-$ \\
		J2145$-$0750 & 11.0 &  & -12.69 & 1.30 \\
		J2317+1439 & 11.0 & & $-$ & $-$ \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}

%We constructed simulated data sets using the same procedure as in \citet{tlb+2017}. 
%The simulated PTAs consisted of 10 MSPs with 
%total observation times and noise properties chosen to be 
%similar to the properties of the EPTA \cite{caballero+2016} (see Table~\ref{tab:sim}). 
%We also injected a stochastic GWB amplitude with $\Agw = 5\times10^{-15}$.
%\begin{table}[htb]
%	\setlength{\tabcolsep}{5pt}
%	\caption{Pulsar parameters used in simulated PTA}
%	\begin{center}
%	\begin{tabular}{ccccc}
%		\hline\hline
%    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
%		\hline
%		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
%		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
%		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
%		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
%		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
%		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
%		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
%		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
%		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
%		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
%    		\hline\hline
%	\end{tabular}
%	\end{center}
%	\label{tab:sim}
%\end{table}

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a simulation with a GWB with $\Agw=5\times10^{-15}$. 
%We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%for a common red process. 
For this particular realization of the GWB, 
the fixed-noise analysis 
using the individual noise results gives 
$\hat{A}^2 = 6.6\times10^{-30}$ with $\mathrm{SNR} = 2.4$, 
and the fixed-noise analysis using the common noise results gives 
$\hat{A}^2 = 2.6\times10^{-29}$ with $\mathrm{SNR} = 6.0$. 
The noise-marginalized analysis gives $\hat{A}^2 = (2.5 \pm 0.1)\times10^{-29}$ 
with $\mathrm{SNR}=4.8\pm0.8$. 
The value of $\hat{A}^2$ from the fixed-noise analysis using the individual noise results 
is significantly lower than the injected level of the GWB, while 
the values of $\hat{A}^2$ from the fixed-noise analysis using the common noise results 
and the noise-marginalized analysis are in good agreement with each other and the injected value. 
The fixed-noise analysis using the individual noise results also gives a significantly lower SNR 
than the analyses using the common noise results. 
The mean SNR from the noise-marginalized analysis is smaller 
than the SNR found with the fixed-noise analysis using the common noise results, 
but the two are consistent at $1.6\sigma$.
\begin{figure}[htb]
	\includegraphics[width=0.95\columnwidth]{plots/optstat_A5e-15_dataset11.pdf}
	\caption{Optimal statistic for a simulated PTA dataset containing an injected GWB. 
			We compare the values found using a fixed-noise analysis 
			with noise parameters taken from 
			individual and common noise analyses 
			(dashed blue lines and solid orange lines, respectively) 
			to the values found by marginalizing over 
			the individual pulsars' red noise parameters (green histograms). 
			The dotted vertical line indicates the injected value, $\Agw^2 = 2.5 \times 10^{-29}$. 
			The fixed-noise analysis 
			using the individual noise results underestimates $\Agw$, 
			while the fixed-noise analysis using the common noise results 
			and the noise-marginalized analysis more accurately recover $\Agw$.}
	\label{fig:os_dataset_sample}
\end{figure}
%\begin{figure}[ht]
%	\includegraphics[width=0.9\columnwidth]{plots/os_dataset50.pdf}
%	\caption{Optimal statistic and SNR for a simulated dataset 
%			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
%			We compare the values found using a fixed-noise analysis (dashed blue line) to the 
%			mean optimal statistic and mean SNR found by marginalizing over 
%			the individual pulsars' red noise parameters (solid orange line). 
%			We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%			for a common red process (solid green line). 
%			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
%			For this particular realization of the stochastic background, the fixed-noise analysis 
%			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
%			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$. 
%			For comparison, the Bayesian analysis gives a mean value of $\Agw^2 = 4.5\times10^{-29}$.}
%	\label{fig:os_dataset_sample}
%\end{figure}

In Fig.~\ref{fig:os_datasetstats} we show the optimal statistic 
for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$ 
computed using the three techniques described above. 
Using the noise values from individual noise analyses 
systematically underestimates the strength of the GWB, 
while using the noise values from a common noise analysis 
more accurately recovers the injected value. 
The fixed-noise analysis using the individual noise results finds 
$\hat{A}^2 = (7.9 \pm 6.8) \times10^{-30}$ and $\rho = 2.3 \pm 1.5$, 
averaging over realizations of the GWB. 
The fixed-noise and noise-marginalized analyses 
using the common noise results both give 
$\hat{A}^2 = (2.4\pm1.2)\times10^{-29}$ and $\rho = 4.1 \pm 1.7$.
\begin{figure}[htb]
	\includegraphics[width=0.95\columnwidth]{plots/optstat_A5e-15.pdf}
	\caption{Optimal statistic and SNR for 200 simulated datasets 
			containing an injected GWB. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			We show the values found by fixing the red noise 
			based on the results from individual and common noise analyses 
			(blue and orange lines, respectively) 
			as well as the mean optimal statistic and mean SNR found by marginalizing over 
			the red noise parameters using the results of a common noise analysis (green lines). 
			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
			that are significantly lower than the injected value, 
			while using the values from a common noise analysis gives values of $\hat{A}^2$ 
			that are closer to the injected value.}
	\label{fig:os_datasetstats}
\end{figure}

The fixed-noise and noise-marginalized analyses give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing datasets 
containing smaller injected values of $\Agw$. 
In Fig.~\ref{fig:os_compare} we show a P--P plot 
of the cumulative fraction of simulations for which the injected $\Agw^2$ lies within 
a given confidence interval of the measured $\hat{A}^2$, 
assuming a Gaussian distribution with mean equal to the measured $\hat{A}^2$ 
and variance $\sigma^2_{\hat{A}^2}$, 
determined from the distribution of $\hat{A}^2$ for our simulations. 
If $\hat{A}^2$ has a Gaussian distribution centered around $\Agw^2$, 
the curves should lie along a straight line with slope 1 (the dotted, diagonal line).

We compare the three methods for computing the optimal statistic for 
simulations with $\Agw=5\times10^{-15}$, $\Agw=10^{-15}$, and $\Agw=5\times10^{-16}$. 
The fixed-noise optimal statistic using the individual noise results systematically underestimates $\hat{A}^2$ 
(Fig.~\ref{fig:os_compare}, left panel). 
The fixed-noise optimal statistic using the common noise results 
estimates $\hat{A}^2$ well for large values of $\Agw$, but for small values it also underestimates $\hat{A}^2$ 
(Fig.~\ref{fig:os_compare}, middle panel). 
The noise-marginalized optimal statistic provides the most accurate estimate of $\hat{A}^2$ 
over the range of \Agw\ considered here (Fig.~\ref{fig:os_compare}, right panel).
\begin{figure*}[t]
	\includegraphics[width=0.95\textwidth]{plots/pp_plot.pdf}
	\caption{P--P plot showing the cumulative fraction of simulations for which $\Agw^2$ lies within 
			a given confidence interval of the measured $\hat{A}^2$. 
			The probability distribution of $\hat{A}^2$ is assumed to be a Gaussian 
			with variance $\sigma^2_{\hat{A}^2}$. 
			The fixed-noise optimal statistic using the individual and common noise results 
			both give biased values of $\hat{A}^2$, particularly for small values of \Agw, 
			while the noise-marginalized optimal statistic gives more accurate values of $\hat{A}^2$ 
			over a large range of injected values of \Agw.}
	\label{fig:os_compare}
\end{figure*}

@@

The optimal statistic is particularly well-suited to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the ORF 
in Eq.~\eqref{eq:}. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we compute the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section in order to determine 
how well we can distinguish a GW background from a monopole or dipole signal. 
For a monopole signal, the ORF becomes simply
$\Gamma(\theta_{IJ}) = 1$, 
while for a dipole signal, the ORF becomes
$\Gamma(\theta_{IJ}) = \cos\theta_{IJ}$.

Our ability to distinguish between different spatial correlations 
depends on the strength of the GWB 
and the angular separations between pulsar pairs, $\theta_{IJ}$. 
We can determine the overlap between ORFs corresponding to different spatial correlations 
by computing the ``match statistic'' \citep{cs2016},
\begin{equation}
	\bar{M} = \frac{\sum_{I,J \neq I} \Gamma_{IJ} \Gamma'_{IJ}}{\sqrt{ \left( \sum_{I, J \neq I} \Gamma_{IJ} \Gamma_{IJ} \right) \left( \sum_{I, J \neq I} \Gamma'_{IJ} \Gamma'_{IJ} \right)}} \,,
\end{equation}
where $I$,$J$ index the pulsar number, and $\Gamma$ and $\Gamma'$ are two different ORFs. 
For the 18 pulsars used in these simulations, the 
match statistic for monopole and Hellings-Downs correlations is $\bar{M} = 0.264$, 
and the match statistic for dipole and Hellings-Downs correlations is $\bar{M} = 0.337$. 
These match statistics describe a fundamental limit on our ability 
to identify the spatial correlations of a common red signal as Hellings-Downs 
rather than monopole or dipole 
that depends only on the number of pulsars in our PTA and their sky positions.

Figure~\ref{fig:os_ORF} shows the noise-marginalized 
mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 200 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
We find a noise-marginalized mean SNR above 1.0 in 97\% of our simulated data sets 
using the quadrupole ORF, and in 50\% and 68\% of our simulated data sets 
using the monopole and dipole ORFs, respectively. 
The mean SNR using the quadrupole ORF, averaged over realizations of the GWB, is 4.1, 
and we find an SNR greater than this using the monopole and dipole ORFs in just 
3\% and 3.5\% of our simulations, respectively.
\begin{figure}[t]
	\includegraphics[width=0.95\columnwidth]{plots/optstat_spatial_A5e-15.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 200 simulated datasets 
			containing an injected GWB with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), 
			and quadrupole (green) spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

>>>>

The optimal statistic is constructed from the autocovariance and cross-covariance matrices 
${\bf{P}}_a$ and ${\bf{S}}_{ab}$,
\begin{eqnarray}
	{\bf{P}}_a &=& \left\langle \res_a \res_a^T \right\rangle \,, \\
	{\bf{S}}_{ab} &=& \left. \left\langle \res_a \res_b^T \right\rangle \right|_{a \neq b} \,.
\end{eqnarray}
where $\res_a$ is a vector of the residuals of the $a$th pulsar in the PTA. 
In the frequency domain, these can be written as
\begin{eqnarray}
	{\bf{P}}_a^{ij} &=& {\bf{R}}_a \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_a(f) \right] {\bf{R}}_a^T \,, \\
	{\bf{S}}_{ab}^{ij} &=& \Gamma_{ab} \; {\bf{R}}_a \left[ \int_{-\infty}^\infty df \; e^{2\pi i f \tau_{ij}} P_\mathrm{gw}(f) \right] {\bf{R}}_b^T \,, \label{eq:Sab}
\end{eqnarray}
where $i,j$ index the matrix elements, $\tau_{ij} \equiv |t_i - t_j|$ is the time interval between 
two observations, 
$\mathbf{R}$ is the timing model fitting matrix, 
$P_\mathrm{gw}(f)$ is the GW power spectrum, 
and $P_a(f)$ is the power spectrum for the $a$th pulsar. 
$\Gamma_{ab}$ is the overlap reduction function (ORF) 
that describes the spatial correlation between two pulsars' residuals \citet{hd1983}, 
\begin{equation}
	\Gamma_{ab} =  \,,
\end{equation}
where $\theta_{ab}$ is the angle between the pulsars. 
The optimal statistic $\hat{A}^2$ is given by
\begin{equation}
	\hat{A}^2 = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) } \,,
\end{equation}
where $\tilde{{\bf{S}}}_{ab}$ is the amplitude-independent cross-correlation matrix,
\begin{equation}
	\Agw^2 \tilde{{\bf{S}}}_{ab} = {\bf{S}}_{ab} \,.
\end{equation}
This definition of the optimal statistic ensures that  
$\langle \hat{A}^2 \rangle = \Agw^2$. 
If $\Agw=0$, the variance of the optimal statistic is
\begin{equation}
	\sigma_0 = \left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{-1/2} \,.
\end{equation}
For a measured value of $\hat{A}^2$, 
the significance of $\hat{A}^2 \neq 0$ is given by the signal-to-noise ratio (SNR) 
\begin{equation}
	\rho = \frac{\hat{A}^2}{\sigma_0} = \frac{\sum_{ab} \res_a^T {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \res_b}{\left[ \sum_{ab} \Tr \left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2}} \,.
\end{equation}
The expectation value of the SNR, averaged over all realizations of the stochastic background, is
\begin{equation}
	\left\langle \rho \right\rangle = \Agw^2 \left[ \sum_{ab} \Tr\left( {\bf{P}}_a^{-1} \tilde{{\bf{S}}}_{ab} {\bf{P}}_b^{-1} \tilde{{\bf{S}}}_{ba} \right) \right]^{1/2} \,.
\end{equation}

@@

We used these three techniques to compute the optimal statistic 
for simulated ``NANOGrav-like'' data sets. 
These consisted of 18 MSPs with observation times up to 11 years 
and sky positions matching the 18 longest-observed pulsars in the 
NANOGrav 11-year data set \citep{abb+17}.
with the observation times increased so that the longest-observed pulsars 
had TOAs spanning over 11 years. 
We include white noise for all pulsars, plus 
red noise parametrized as a power law, $P_a(f) = A_\mathrm{red} f^{-\gamma}$, 
for those pulsars that show evidence of red noise 
(see Table~\ref{tab:sim} for more details).

\begin{table}[tb]
	\setlength{\tabcolsep}{5pt}
	\caption{Pulsar parameters used in simulated PTAs. \sv{finish filling in this table}}
	\begin{center}
	\begin{tabular}{ccccc}
		\hline\hline
    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
		\hline
		J0030+0451 & 11.0 & & -13.93 & 3.56 \\
		J0613$-$0200 & 11.0 &  & -13.14 & 1.22 \\
		J1012+5307 & 11.0 &  & -12.79 & 1.51 \\
		J1024$-$0719 & 6.0 & & $-$ & $-$ \\
		J1455$-$3330 & 11.0 & & $-$ & $-$ \\
		J1600$-$3053 & 8.0 & & $-$ & $-$ \\
		J1614$-$2230 & 7.0 & & $-$ & $-$ \\
		J1640+2224 & 11.0 & & $-$ & $-$ \\
		J1713+0747 & 11.0 & & -14.14 & 1.58 \\
		J1741+1351 & 6.0 & & $-$ & $-$ \\
		J1744$-$1134 & 11.0 & & $-$ & $-$ \\
		B1855+09 & 11.0 & & -13.75 & 3.54 \\
		J1853+1303 & 7.0 & & $-$ & $-$ \\
		J1909$-$3744 & 11.0 &  & -13.84 & 1.74  \\
		J1918$-$0642 & 11.0 & & $-$ & $-$ \\
		J2010$-$1323 & 7.0 & & $-$ & $-$ \\
		J2145$-$0750 & 11.0 &  & -12.69 & 1.30 \\
		J2317+1439 & 11.0 & & $-$ & $-$ \\
    		\hline\hline
	\end{tabular}
	\end{center}
	\label{tab:sim}
\end{table}
%\begin{table}[htb]
%	\setlength{\tabcolsep}{5pt}
%	\caption{Pulsar parameters used in simulated PTA}
%	\begin{center}
%	\begin{tabular}{ccccc}
%		\hline\hline
%    		Pulsar	& $T_\mathrm{obs}$ (yrs) & $\sigma_w$ ($\mu\mathrm{s}$) & $\log_{10}A_\mathrm{red}$ & $\gamma_\mathrm{red}$ \\
%		\hline
%		J0613$-$0200 & 16.054 & 1.58 & -13.90 & 3.18 \\
%		J0751+1807 & 17.606 & 2.60 & -14.14 & 2.58 \\
%		J1012+5307 & 16.831 & 1.47 & -13.09 & 1.65 \\
%		J1640+2224 & 16.735 & 1.99 & -13.24 & 0.03 \\
%		J1643$-$1224 & 17.300 & 1.65 & -18.56 & 4.04 \\
%		J1713+0747 & 17.657 & 0.26 & -14.90 & 4.85 \\
%		J1744$-$1134 & 17.250 & 0.65 & -13.60 & 2.00 \\
%		J1857+0943 & 17.310 & 1.51 & -16.00 & 1.35 \\
%		J1909$-$3744 & 9.379 & 0.12 & -13.99 & 2.06 \\
%		J2145$-$0750 & 17.161 & 1.19 & -13.87 & 4.02 \\
%    		\hline\hline
%	\end{tabular}
%	\end{center}
%	\label{tab:sim}
%\end{table}

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a simulation with a GWB with $\Agw=5\times10^{-15}$. 
%We also include the distribution of $\Agw^2$ found using the results of a Bayesian analysis 
%for a common red process. 
For this particular realization of the GWB, 
the fixed-noise analysis 
using the individual noise results gives 
$\hat{A}^2 = 6.6\times10^{-30}$ with $\mathrm{SNR} = 2.4$, 
and the fixed-noise analysis using the common noise results gives 
$\hat{A}^2 = 2.6\times10^{-29}$ with $\mathrm{SNR} = 6.0$. 
The noise-marginalized analysis gives $\hat{A}^2 = (2.5 \pm 0.1)\times10^{-29}$ 
with $\mathrm{SNR}=4.8\pm0.8$. 
The value of $\hat{A}^2$ from the fixed-noise analysis using the individual noise results 
is significantly lower than the injected level of the GWB, while 
the values of $\hat{A}^2$ from the fixed-noise analysis using the common noise results 
and the noise-marginalized analysis are in good agreement with each other and the injected value. 
The fixed-noise analysis using the individual noise results also gives a significantly lower SNR 
than the analyses using the common noise results. 
The mean SNR from the noise-marginalized analysis is smaller 
than the SNR found with the fixed-noise analysis using the common noise results, 
but the two are consistent at $1.6\sigma$.

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15_dataset11.pdf}
	\caption{Optimal statistic for a simulated PTA dataset containing an injected GWB. 
			We compare the values found using a fixed-noise analysis 
			with noise parameters taken from 
			individual and common noise analyses 
			(dashed blue lines and solid orange lines, respectively) 
			to the values found by marginalizing over 
			the individual pulsars' red noise parameters (green histograms). 
			The dotted vertical line indicates the injected value, $\Agw^2 = 2.5 \times 10^{-29}$. 
			The fixed-noise analysis 
			using the individual noise results underestimates $\Agw$, 
			while the fixed-noise analysis using the common noise results 
			and the noise-marginalized analysis more accurately recover $\Agw$.}
	\label{fig:os_dataset_sample}
\end{figure}

In Fig.~\ref{fig:os_datasetstats} we show the optimal statistic 
for 200 different realizations of a GWB with $\Agw = 5\times10^{-15}$ 
computed using the three techniques described above. 
Using the noise values from individual noise analyses 
systematically underestimates the strength of the GWB, 
while using the noise values from a common noise analysis 
more accurately recovers the injected value. 
The fixed-noise analysis using the individual noise results finds 
$\hat{A}^2 = (7.9 \pm 6.8) \times10^{-30}$ and $\rho = 2.3 \pm 1.5$, 
averaging over realizations of the GWB. 
The fixed-noise and noise-marginalized analyses 
using the common noise results both give 
$\hat{A}^2 = (2.4\pm1.2)\times10^{-29}$ and $\rho = 4.1 \pm 1.7$.

\begin{figure}[tb]
	\includegraphics[width=0.9\columnwidth]{plots/optstat_A5e-15.pdf}
	\caption{Optimal statistic and SNR for 200 simulated datasets 
			containing an injected GWB. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			We show the values found by fixing the red noise 
			based on the results from individual and common noise analyses 
			(blue and orange lines, respectively) 
			as well as the mean optimal statistic and mean SNR found by marginalizing over 
			the red noise parameters using the results of a common noise analysis (green lines). 
			Using the red noise values from individual noise analyses leads to values of $\hat{A}^2$ 
			that are significantly lower than the injected value, 
			while using the values from a common noise analysis gives values of $\hat{A}^2$ 
			that are closer to the injected value.}
	\label{fig:os_datasetstats}
\end{figure}

\begin{figure*}[ht]
	\includegraphics[width=0.95\textwidth]{plots/pp_plot.pdf}
	\caption{P--P plot showing the cumulative fraction of simulations for which $\Agw^2$ lies within 
			a given confidence interval of the measured $\hat{A}^2$. 
			The probability distribution of $\hat{A}^2$ is assumed to be a Gaussian 
			with variance $\sigma^2_{\hat{A}^2}$. 
			The fixed-noise optimal statistic using the individual and common noise results 
			both give biased values of $\hat{A}^2$, particularly for small values of \Agw, 
			while the noise-marginalized optimal statistic gives more accurate values of $\hat{A}^2$ 
			over a large range of injected values of \Agw.}
	\label{fig:os_compare}
\end{figure*}

The fixed-noise and noise-marginalized analyses give the same results 
for $\Agw = 5\times10^{-15}$, 
but there are differences between them when analyzing datasets 
containing smaller injected values of $\Agw$. 
In Fig.~\ref{fig:os_compare} we show a P--P plot 
of the cumulative fraction of simulations for which the injected $\Agw^2$ lies within 
a given confidence interval of the measured $\hat{A}^2$, 
assuming a Gaussian distribution with mean equal to the measured $\hat{A}^2$ 
and variance $\sigma^2_{\hat{A}^2}$, 
determined from the distribution of $\hat{A}^2$ for our simulations. 
If $\hat{A}^2$ has a Gaussian distribution centered around $\Agw^2$, 
the curves should lie along a straight line with slope 1 (the dotted, diagonal lines in Fig.~\ref{fig:os_compare}).

We compare the three methods for computing the optimal statistic for 
simulations with $\Agw=5\times10^{-15}$, $\Agw=10^{-15}$, and $\Agw=5\times10^{-16}$. 
The fixed-noise optimal statistic using the individual noise results systematically underestimates $\hat{A}^2$ 
(Fig.~\ref{fig:os_compare}, left panel). 
The fixed-noise optimal statistic using the common noise results 
estimates $\hat{A}^2$ well for large values of $\Agw$, but for small values it also underestimates $\hat{A}^2$ 
(Fig.~\ref{fig:os_compare}, middle panel). 
The noise-marginalized optimal statistic provides the most accurate estimate of $\hat{A}^2$ 
over the range of \Agw\ considered here (Fig.~\ref{fig:os_compare}, right panel).

@@

The optimal statistic is particularly well-suited to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the ORF 
in Eq.~\eqref{eq:Sab}. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we compute the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section in order to determine 
how well we can distinguish a GW background from a monopole or dipole signal. 
For a monopole signal, the ORF becomes simply
$\Gamma(\theta_{ab}) = 1$, 
while for a dipole signal, the ORF becomes
$\Gamma(\theta_{ab}) = \cos\theta_{ab}$.

Our ability to distinguish between different spatial correlations 
depends on the strength of the GWB 
and the angular separations between pulsar pairs, $\theta_{ab}$. 
We can determine the overlap between ORFs corresponding to different spatial correlations 
by computing the ``match statistic'' \citep{cs2016},
\begin{equation}
	\bar{M} = \frac{\sum_{a,b \neq a} \Gamma_{ab} \Gamma'_{ab}}{\sqrt{ \left( \sum_{a, b \neq a} \Gamma_{ab} \Gamma_{ab} \right) \left( \sum_{a, b \neq a} \Gamma'_{ab} \Gamma'_{ab} \right)}} \,,
\end{equation}
where $\Gamma$ and $\Gamma'$ are two different ORFs. 
For the 18 pulsars used in these simulations, the 
match statistic for monopole and Hellings-Downs correlations is $\bar{M} = 0.264$, 
and the match statistic for dipole and Hellings-Downs correlations is $\bar{M} = 0.337$. 
These match statistics describe a fundamental limit on our ability 
to identify the spatial correlations of a common red signal as Hellings-Downs 
rather than monopole or dipole 
that depends only on the number of pulsars in our PTA and their sky positions.
